---
title: "STAT1379/COMP6179 — Section 1 Report"
author: "Mohammed Saif Ur Rahman (48941298)"
params:
  data_file: "inst/extdata/project2025.csv"
  alpha: 0.05
format:
  html: default
  pdf: default
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
link-citations: true
---

```{r}
library(here)
library(readr)
library(dplyr)
library(ggplot2)
library(gt)        
library(broom)     
library(project2025)  
```

## Abstract

This report analyses anthropometric and behavioural data from 1,000 adults aged 26–45 to explore three research questions: (1) whether a linear relationship exists between height and weight, (2) whether the average height differs between males and females, and (3) whether gender and physical activity levels are associated. Using R and Quarto for reproducible analysis, we apply linear regression (lm()), a two-sample t-test assuming equal variances, and a chi-square test of independence. Results show a statistically significant positive relationship between height and weight (p\<0.05), supporting a linear association. The t-test indicates that mean height differs by gender (p\<0.05), while the chi-square test suggests gender and physical activity are largely independent (p\>0.05). These findings illustrate how classical inference techniques can provide evidence-based insights into human physical and lifestyle characteristics.

## Introduction

This report addresses three questions using a dataset of adults aged 26–45: (i) linear association between Height and Weight; (ii) equality of mean Height across Gender (Male vs Female); and (iii) association between Gender and Physical Activity (None/Moderate/Intense). We use standard inferential tools and a reproducible Quarto workflow [@ggplot2; @r4ds].

## Methodology

**Q1 (Linear regression).**\
We model the relationship between weight and height as:

$$
\text{Weight}_i = \beta_0 + \beta_1 \text{Height}_i + \varepsilon_i,
\qquad
H_0: \beta_1 = 0,\;
H_1: \beta_1 \neq 0.
$$

**Q2 (Two-sample t-test, equal variances).**\
Let $\mu_M$ and $\mu_F$ represent mean heights for males and females.

$$
H_0: \mu_M = \mu_F, \qquad
H_1: \mu_M \neq \mu_F, \quad
\alpha = 0.05.
$$

**Q3 (Chi-square test of independence).**\
With observed counts $O_{ij}$ and expected counts $E_{ij}$:

$$
H_0: \text{Gender} \perp \text{Physical Activity}, \qquad
\chi^2 = \sum_{i,j} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}.
$$

We report the test statistic, degrees of freedom, $p$-value, and a decision at $\alpha = 0.05$.

## Results

```{r}
# load package and data
library(project2025)   # my package name
library(readr)

dat <- readr::read_csv(here::here(params$data_file)) |>
dplyr::rename(
ID = ID,
Gender = gender,
Height = height,
Weight = weight,
PhysicalActivity = phys
)


res1 <- height_weight_lm(dat)
res2 <- gender_height_ttest(dat)
res3 <- gender_activity_chisq(dat)

# auto-updating textual decisions
res1; res2; res3
# summary of height by gender (for nice descriptive table)

by_gender <- dat |>
dplyr::group_by(Gender) |>
dplyr::summarise(
n = dplyr::n(),
mean_height = mean(Height),
sd_height = sd(Height),
.groups = "drop"
)

gt(by_gender) |>
fmt_number(columns = c(mean_height, sd_height), decimals = 2) |>
tab_header(
title = md("**Height Summary by Gender**"),
subtitle = paste0("n = ", nrow(dat))
)

```

### Plots

### Q1: Height–Weight (Linear Model)

```{r}
res1$plots$scatter_lm
```

**Q1 decision (auto):**  
Slope = `r round(res1$statistic["slope.Height"], 3)`, R² = `r round(res1$statistic["R2"], 3)`, *p* = `r signif(res1$p_value, 4)`.

Decision at α = `r params$alpha`: **`r ifelse(res1$p_value < params$alpha, "Reject H0 (evidence of linear association).", "Fail to reject H0 (no clear linear association).")`**.  
`r res1$conclusion`

### Q2: Mean Height by Gender (t-test)

```{r}
res2$plots$box

```

**Q2 decision (auto):**  
*t* (df = `r round(res2$statistic["df"], 1)`) = `r round(res2$statistic["t"], 3)`, *p* = `r signif(res2$p_value, 4)`.

Male mean = `r round(res2$estimates["male"], 2)` cm; Female mean = `r round(res2$estimates["female"], 2)` cm; Δ (M − F) = `r round(res2$estimates["diff_male_female"], 2)` cm.

Decision at α = `r params$alpha`: **`r ifelse(res2$p_value < params$alpha, "Reject H0 (means differ).", "Fail to reject H0 (means similar).")`**.  
`r res2$conclusion`


### Q3: Gender × Physical Activity (Chi-square)

```{r}
res3$plots$stack100

```

**Q3 decision (auto):**  
χ² (df = `r res3$statistic["df"]`) = `r round(res3$statistic["chi_square"], 2)`, *p* = `r signif(res3$p_value, 4)`.

Decision at α = `r params$alpha`: **`r ifelse(res3$p_value < params$alpha, "Reject H0 (associated).", "Fail to reject H0 (independent).")`**.  
`r res3$conclusion`


## Conclusion

We examined three questions on adults aged 26–45. For the height–weight relationship, the estimated slope was `r round(res1$statistic["slope.Height"],3)` with R² = `r round(res1$statistic["R2"],3)`, leading us to **`r ifelse(res1$p_value < params$alpha, "reject", "not reject")`** the null of zero slope at α = `r params$alpha`. The two-sample *t*-test indicated male and female mean heights were **`r ifelse(res2$p_value < params$alpha, "different", "not detectably different")`**. The chi-square test suggested gender and physical activity were **`r ifelse(res3$p_value < params$alpha, "associated", "independent")`**. Conclusions depend on standard assumptions (equal variances for the *t*-test; adequate expected counts for chi-square; basic regression diagnostics).
